Clinical prediction with localized modeling using similarity-based cohorts: A scoping review

### 2024-06-17 docs/localized-modeling-medRxiv/localized-modeling-medRxiv.Rmd
prepare arxiv manuscript

---
# Supported options: 
#   sn-nature:       Style for submissions to Nature Portfolio journals
#   sn-basic:        Basic Springer Nature Reference Style/Chemistry Reference Style
#   sn-mathphys:     Math and Physical Sciences Reference Style
#   sn-aps:          American Physical Society (APS) Reference Style
#   sn-vancouver:    Vancouver Reference Style
#   sn-apa:          APA Reference Style 
#   sn-chicago:      Chicago-based Humanities Reference Style
#   default:         Default

classoptions: 
  - sn-nature      
  # - Numbered      # Optional: Use numbered references instead of namedate references (only for sn-nature, sn-basic, sn-vancouver, sn-chicago, sn-mathphys or sn-nature)
  # - referee       # Optional: Use double line spacing 
  # - lineno        # Optional: Add line numbers
  # - iicol         # Optional: Double column layour

title: "Clinical prediction with localized modeling using similarity-based cohorts: A scoping review"
titlerunning: Clinical Prediction with Localized Modeling
authors: 
  - firstname: Adi
    lastname: Cohen
    degrees: BS
    email: ac4569@mynsu.nova.edu
    affiliation: 1
  - firstname: Patti
    lastname: McCall-Junkin
    degrees: MA, MLS
    email: pattimccall.junkin@gmail.com
    affiliation: 2
  - firstname: Jason Cory
    lastname: Brunson
    degrees: PhD
    email: jason.brunson@medicine.ufl.edu
    orcid: 0000-0003-3126-9494
    affiliation: 3
    corresponding: TRUE
affiliations:
  - number: 1
    info:
      orgdiv: College of Osteopathic Medicine
      orgname: Nova Southeastern University
    address:
        street: 3200 S University Drive
        city: Davie
        postcode: 33328
        state: FL
        country: USA
  - number: 2
    info:
      orgdiv: Smathers Libraries Academic Research & Consulting Services
      orgname: University of Florida
    address:
        street: P.O. Box 117000
        city: Gainesville
        postcode: 32611-7000
        state: FL
        country: USA
  - number: 3
    corresponding: TRUE
    info:
      orgdiv: Laboratory for Systems Medicine
      orgname: University of Florida
    address:
        street: P.O. Box 100225 JHMHC
        city: Gainesville
        postcode: 32610-0225
        state: FL
        country: USA
keywords:
  - case-based reasoning
  - localized modeling
  - nearest neighbors
  - clinical decision support
  - scoping review
# pacs: 
#   jel:
#     - "D8"
#     - "H51"
#   msc:
#     - "35A01"
#     - "65L10"

# Sample for structured abstract
abstract: |
  Background:
  Applications of classical case-based reasoning (CBR) have given rise to a family of techniques we call “localized models”, in which a statistical model is fitted to a neighborhood of \hl{labeled cases matched by similarity to a target case}. We aim to describe clinical and health applications of localized models to date and propose a general framework for their design and evaluation.

  Methods:
  We searched four bibliographic platforms during 2021 July 19–22\hl{, updated 2024 January 24}. We set four eligibility criteria to identify applications of localized models to clinical and health tasks. Two authors divided title/abstract screening and reviewed screened entries for inclusion. We discussed settings, tasks, and tools; identified and tabulated themes; and synthesized the methods into a general framework.

  Results:
  Of \hl{1,657} search results, \hl{360} were reviewed, then combined with 43 publications that seeded the review and 1 obtained by citation tracking. \hl{27} were included, published 1997–\hl{2022}. The specificity of search terms was poor, and inter-rater reliability was low. Almost all models were predictive, the most common tasks being prognosis and diagnosis. Most studies used clinical, occasionally laboratory and image, data. Several addressed memory and runtime costs. A general technique that specializes to most of those reviewed \hl{involved matching, retrieval, fitting, and evaluation steps that could optionally be supervised, optimized, or recursively performed}.

  Conclusions:
  Localized models have potential to improve the performance of clinical decision support tools while maintaining interpretability, \hl{but rigorous comparisons to competing methods must be conducted and} computational hurdles \hl{must be} \hl{overcome.} We hope that our review will spur future work on efficiency, reproducibility, and user needs.

bibliography:
  - SR-bibliography.bib
  - SR-reviewers.bib
  - SR-0-Found.bib
  - SR-2-Included.bib
  - SR-3-Tracked.bib
  - SR-5-Screened.bib

header-includes: |
  \usepackage{lscape}
  \newcommand{\landscapebegin}{\begin{landscape}}
  \newcommand{\landscapeend}{\end{landscape}}

  \usepackage{soul}
  %%\definecolor{hlcolor}{rgb}{.80,.90,1}
  \definecolor{hlcolor}{rgb}{1,1,1}
  \sethlcolor{hlcolor}

  %% Per the spinger doc, new theorem styles can be included using built in style, 
  %% but it seems the don't work so commented below
  %\theoremstyle{thmstyleone}%
  \newtheorem{theorem}{Theorem}%  meant for continuous numbers
  %%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
  %% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
  \newtheorem{proposition}[theorem]{Proposition}%
  %%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.

  %% \theoremstyle{thmstyletwo}%
  \theoremstyle{remark}
  \newtheorem{example}{Example}%
  \newtheorem{remark}{Remark}%

  %% \theoremstyle{thmstylethree}%
  \theoremstyle{definition}
  \newtheorem{definition}{Definition}%

output:
  rticles::springer_article:
    keep_tex: true
    latex_engine: xelatex
    # citation_package: natbib
    citation_package: "default"
---

@@

A main driver of the development of clinical information systems (CIS) and clinical decision support (CDS) tools has been to leverage the advantages of routinely collected health data toward improving care.
These advantages include their immediate accessibility through the institutional EHR, billing database, or other source; their specificity to the institution and the population it serves; and the closeness in time of the available data.
Though by definition not collected for research use, the secondary research use of routinely collected health data has been put forth, \hl{as} "practice-based evidence", \hl{a} complement to the paradigm of evidence-based practice.

Over the same period of time, interest in individualizing care from population-derived evidence-based recommendations to specific patient needs has driven the application of advanced computational tools, including artificial intelligence (AI). Whereas classical models produced formulae involving a limited set of data elements that would be uniformly applied to all patients, AI models often process hundreds or more variables in opaque ways to yield predictions that depend on not only their values but their associations and \hl{interplay with each other}. The loss of straightforward interpretations of these models has limited their practical uptake and motivated the construction of explanatory statistics for opaque models as well as the development of more interpretable complex models.

One of few methodologies to address all three of these established needs is one we call _localized modeling_, which appears to have been introduced several times independently under different names, and in slightly different forms.
The approach is a specialized form of case-based reasoning (CBR) that relies on a patient similarity measure[^psm] to extract a cohort of past or training cases that then inform the diagnosis, prognosis, or care of a new or test case.
\hl{Traditional} CBR returns these retrieved cases to inform human decision-making\hl{;} the related nearest neighbors (NN) technique generates predictions from cohorts automatically via averaging (regression) or voting (classification) of their outcomes.
In contrast, localized modeling fits families of predictive models, \hl{for example} generalized linear models (GLMs), \hl{to retrieved} similarity cohorts in order to generate predictions.
Localized models thus provide \hl{an individualized way to fit fully interpretable models to large population data}.

@@

This approach harkens to the aspirational "green button" that would, in response to a query, automatically retrieve patient data from an institution’s records with which to conduct on-demand retrospective studies for an individual patient [@Longhurst2014].
\hl{Several hurdles face the deployment of such an approach in practice, but its benefits must first be demonstrated.}
Our motivations in this scoping review are to describe the settings in and problems with which this approach has been tasked, to attempt to organize them within a common framework, and to evaluate the promise they show toward achieving this or other needs in clinical informatics.
Along the way, we attempt to reconcile terminology, summarize motivations and evaluations, and propose valuable follow-up work.

@@

@Kolodner1992 distinguished two styles of CBR: problem-solving, which is more procedural and used when objectives are more clearly defined, and interpretative, which provides categorizations and justifications for possible solutions. A similar dichotomy is commonly used to distinguish the performance criterion for the usefulness of predictive models from the interpretability criterion for their usability. In these terms, localized models are highly procedural: The step of fitting a predictive model to a retrieved cohort \hl{is} an automated adaptive strategy [@Begum2011], and the parameters that govern cohort retrieval can be tuned alongside model hyperparameters in a \hl{machine learning (ML)} workflow. Accordingly, localized models have been primarily used for and evaluated on their ability to predict classes or outcomes.
Nevertheless, as some recent applications have \hl{shown}, they can be used to draw inferences about the importance of different risk factors to specific individuals.
While most ML models come equipped with measures of feature importance and model-agnostic tools can generate explanations for model predictions, these quantifications are not directly interpretable model components analogous to the split nodes of a decision tree (DT) or the \hl{coefficients} of a GLM.
While the patient similarity measure used to retrieve each cohort may be complicated, the cohort itself can be directly inspected by the user. Provided the model family fitted to the cohorts is interpretable, the localized model \hl{inherits} this property. Thus, localized models may embody both styles of CBR.

@@

Our goals in this review are (1) to describe \hl{applications} of similarity-based localized models using patient-level health data and (2) to provide a general framework for the design and evaluation of such localized models.
\hl{We focus narrowly on localized models, rather than broadly on patient similarity--based clinical decision support tools, so that we may thoroughly assess their value in terms of reported evaluations and comparisons to other methods.
This also allows us to devise a high-level framework that specializes to the majority of these methods, which we then use to frame our discussion and recommendations.}

@@

We expected the general framework to come down to three choices: A patient similarity measure, a cohort selection process, and a statistical model family\hl{. }A general implementation based on this method would allow researchers to expedite every step of the analysis process, including \hl{retrieval}, optimization, and evaluation, and enable sensitivity, robustness, and multiverse analyses that help identify the most consequential choices along the way.

@@

\hl{We repeated the search on 2024 January 24 on the PubMed and Web of Science databases to bring the results up to date.}

@@

Within Covidence, we screened titles and abstracts for relevance. We decided on four eligibility criteria to expedite the screening process.
Articles must be written in English, for readability; they must be original studies, to exclude secondary sources with duplicate information; their use settings must be medical, clinical, or related, to keep our review topical; and it must not be clear that their use of our search terms was different from our intended meaning.
Each of two authors (AC and JCB) screened roughly half of the entries. They regularly reviewed each other’s decisions to improve consistency. In cases of uncertainty, entries were included.
\hl{For the update, one author (PM) screened all results for three criteria and another (JCB) screened the survivors for the fourth criterion.}

@@

Finally, one author (JCB) applied the same review process to the seed set of 43 studies that inspired the review.
\hl{The same author later reviewed the updated results.}

@@

\hl{We next} collected characteristics of included studies. The features included bibliographic fields (date of publication, journal, authors, title, keywords, DOI), study goals (objective, generalizable knowledge, evaluation, clinical/medical domain), data sets (data source, type of data, range of data, number of cases/incidences, number of predictors/features), and methodological choices (types of similarity measure, families of adaptation step/statistical model, method(s) compared against, performance measures, results of evaluations and comparisons, name given to modeling approach). We used these data to detect and visualize trends amongst the included studies.

@@

Rather than an evidence synthesis characteristic of most systematic reviews, we here pursue a methodology synthesis to harmonize largely independent research efforts that have converged on a common \hl{family of techniques}. The goal will be to describe a unified framework for localized models that can be used to guide future study designs and implementations as well as more systematically evaluate variations on the theme and measure the dependence of results on modeling choices.

@@

Figure \ref{fig:prisma} depicts our identification of studies via databases and registers. Following the completion and input of each search string, there were a total of \hl{1,817} sources within all of the platforms used. De-duplication resulted in \hl{1,657} entries, which were added to the title/abstract screening for review. Of these, \hl{360} entries met the screening criteria and were assessed through full text review. Of these, \hl{51} fit the original criteria, and \hl{21} were included as distinct from NN prediction.

```{r prisma, fig.align='center', out.width="80%", echo=FALSE}
#| fig.cap =
#| "PRISMA-S flow chart.
#|  Lowercase letters refer to items obtained from
#|  the seed set (m),
#|  the structured search (n), and
#|  citation/reference tracking (s).
#|  \\hl{Within the structured search results,
#|  summands correspond to original and update.}"
knitr::include_graphics("Fig1.pdf")
```

@@

We then reviewed 43 studies comprising a seed set that inspired this review. After removing duplicates and screening for eligibility, we were left with 6 additional studies [@Park2006; @Lowsky2013; @Lee2015; @Ng2015; @Lee2017; @Wang2019], 1 of which was excluded from the synthesis for using NN prediction. \hl{Reference} tracking from the \hl{43 + 360 = 403} studies assessed for eligibility led us to identify 1 additional study that met our criteria [@Kasabov2010]. This left us with \hl{21 + 1 + 5 = 27} studies included in the review and synthesis.

@@

The `r nrow(tab_synthesis)` included studies were analyzed based on several characteristics, and we report and describe some observations here (Table \ref{tab:synthesis}). The studies were published in a variety of journals, with some of greater frequency, though no single journal published more than 3. The journal _Evolving Systems_ published 2 of the included studies, _Artificial Intelligence in Medicine_ published 3, _Hindawi - Journal of Healthcare Engineering_ published 3\hl{, \emph{Evolving Systems} published 2}, and _PLOS One_ published 2.

Another interesting pattern lay in the years of publication (Figure \ref{fig:year}). While \hl{these studies} trace back to the late 1990s, they have become more common, which suggests that this is an active, though not rapidly expanding, \hl{approach}. We note \hl{that} CBR in health and medicine originated as early as 1990, \hl{and localized modeling emerged soon after} CBR had established itself\hl{;} the idea has been "in the air" for as long as CBR has been in use.

@@

In addition to the aim of its analysis, we coded several aspects of the design of each study, including the source and type of data and the clinical task the model performed (Table \ref{tab:synthesis}), as well as the specific methodology and terminology adopted (Table \ref{tab:composite}).
In the following subsections, we take a closer look at these design elements \hl{and their reported justifications and limitations}.

@@

The most common clinical motivations were individualized detection or diagnosis, prognosis or outcome prediction, and treatment or care recommendation. The plurality focused on prognosis or outcome prediction, often using time-to-event analysis: @Mariuzzi1997 used CBR to predict survival time from several geometric properties of breast tumors. @Lowsky2013 used CBR with non-parametric survival models on registry data to predict patient–graft survival times following kidney transplantation. @Lee2015 and @Lee2017 used localized logistic regression and random forest modeling to predict 30-day mortality following discharge for ICU patients. @Vilhena2016 developed a CBR cycle around a clustering-informed similarity matching procedure and an artificial neural network–based classifier to identify thrombophilia patients at high risk of thrombotic episodes. @Ma2020 took a similarity cohort–based approach to predicting length of stay for ICU patients. @Doborjeh2022 \hl{localized spiking neural networks in order to assess stroke risk from up to 7-day time series of clinical and environmental factors.} @Liu2022 \hl{incorporated global-to-local transfer learning into localized models of acute kidney injury risk in hospitalized patients.} Also of note, from a public health perspective, @Xu2008 used CBR to predict rehabilitation time as well as disability risk for unemployed workers experiencing chronic pain.

@@

The other frequent advantage attributed to localized modeling was interpretability.
@Elter2007 and @Nicolas2014 emphasized the value of the detailed past cases, available to the user, on which predictions are based. This "self-explanation capability" made outputs more intelligible to physicians in the CDS setting.
@Ng2015 additionally pointed out that localized GLMs yield localized effect estimates, which may help investigators identify individually relevant risk factors.
@Liu2022 \hl{took this idea further, systematically comparing regression coefficients for specific risk factors across dozens of diagnostic subpopulations.}
@Doborjeh2022 \hl{used localized time series models to identify environmental changes associated with increased stroke risk.}

@@

CBR can be understood as the opposite side of a trade-off with rule-based reasoning between model size and model complexity\hl{:} @Doyle2006 describe their approach as "knowledge-light", in that "the cases do not contain explicit explanation structures; instead, explanation is achieved by comparison of the query case with retrieved cases". This means that the greatest performance and efficiency challenges in CBR have to do with the retrieval and revision phases in the R4 cycle. Several studies addressed these challenges: @Park2006, while excluded from the synthesis, were the earliest to propose that cohorts be bounded by a similarity threshold rather than by \hl{the} number of cases, \hl{which} improved performance in their experiments. @CampilloGimenez2013 leveraged predictor weights obtained from logistic regression to inform the similarity calculation used in retrieval. @Lowsky2013 used non-parametric models to reduce the computational burden of revision (prediction). @Ma2020 proposed to improve efficiency along the entire R4 cycle, but in particular by only executing task-dependent steps in real time ("just-in-time learning", JITL), and @Ng2021 and @Tang2021 partitioned multiple phases in the R4 cycle into offline and online components, only the latter of which would be performed in real time as new data are received.
@Liu2022 \hl{used transfer learning from globally-fitted models to improve the efficiency of localized models.}

Other studies addressed limitations of available tools. @Lopez2011 implemented a comprehensive CBR tool in response to the lack of general-purpose software, to enable coupling with other tools as well as to expedite development, experimentation, and uptake. Several other teams also set out to develop more \hl{generalizable and data-agnostic} clinical support tools [@Elter2007; @Liang2015; @Ng2015; @Zhang2018].
\hl{However, more informatical and legal challenges to implementation, including interoperability of systems and data and regulations concerning security and privacy, were rarely addressed.
We discuss this further in Section} \ref{feasibility}\hl{.}

@@

Most included studies quantitatively compared the predictive performance of their proposed method(s) to one or more comparators. What we took to be the signature results are collated in Table \ref{tab:performance} in the Appendix. Note that we exercised some judgment in classifying methods as proposals and comparators, as in some cases all methods were original but only some showcased main ideas.
It would be impractical to meta-analyze these numbers due to the great variety of settings, problems, data types\hl{, techniques}, and choices involved.

We do observe one clear pattern: All of the proposed methods that most evidently outperformed their comparators\hl{---}Kohonen + CBR [@Wyns2004], TWNFI [@Song2006, @Kasabov2010], gravitational search algorithms (GSA) [@Liang2015], CBR + rules (+ DML) [@Nicolas2014], Gaussian process regression [@Zhang2018], JITL-ELM [@Ma2020]\hl{---}are hybrids of localized modeling (in some cases CBR) with other techniques, often DML. Though @CampilloGimenez2013 and @Ng2015 report non-superior performance by such hybrids, the pattern suggests the importance of the similarity measure to the retrieval step.
Meanwhile, when proposed approaches targeted cohort demarcation or choice of predictive model\hl{---}statistical CBR [@Park2006]; individualized logistic regression, decision tree, and random forest [@Lee2015; @Lee2017]\hl{---}they did not consistently outperform comparators.

@@

Consistent with their orientation, almost all included studies were conducted using clinical data, only occasionally together with patient-reported (2), \hl{laboratory (3), image (1), and environmental (1)} data. We suggest three reasons for this: First, this literature traces back to the 1990s, before -omic data could be generated cost-effectively at scale. Second, CBR in particular has a strong tradition in clinical decision support, where the focus of our sample remains throughout the review period. Third, because most -omic data are highly homogeneous—all measurements are made along or can be transformed to a common scale, e.g. greyscale pixellations for X-ray images and transcripts per million for RNA-seq data—more deeply theoretical analysis techniques have been developed and come into wide use. While variations on correlation-based approaches like EHR-based phenome-wide association studies and similarity-based methods like CBR itself have been developed, more mechanistic and probabilistic tools have not become a domain standard.

@@

```{r framework, fig.align='center', out.width="100%", echo=FALSE}
#| fig.cap =
#| "General framework for localized models.
#|  Dashed arrows indicate optional steps."
knitr::include_graphics("Fig4.pdf")
```

>>>>

---
title: "Clinical prediction with localized modeling using similarity-based cohorts: A scoping review"
authors: 
  - name: Adi Cohen
    email: ac4569@mynsu.nova.edu
    affiliation: College of Osteopathic Medicine, Nova Southeastern University
    location: Davie, FL
  - name: Patti McCall-Junkin
    email: pattimccall.junkin@gmail.com
    affiliation: Smathers Libraries Academic Research & Consulting Services, University of Florida
    location: Gainesville, FL
  - name: Jason Cory Brunson
    email: jason.brunson@medicine.ufl.edu
    affiliation: Laboratory for Systems Medicine, University of Florida
    location: Gainesville, FL
abstract: |
  Background:
  Applications of classical case-based reasoning (CBR) have given rise to a family of techniques we call “localized models”, in which a statistical model is fitted to a neighborhood of labeled cases matched by similarity to a target case. We aim to describe clinical and health applications of localized models to date and propose a general framework for their design and evaluation.

  Methods:
  We searched four bibliographic platforms during 2021 July 19–22, updated 2024 January 24. We set four eligibility criteria to identify applications of localized models to clinical and health tasks. Two authors divided title/abstract screening and reviewed screened entries for inclusion. We discussed settings, tasks, and tools; identified and tabulated themes; and synthesized the methods into a general framework.

  Results:
  Of 1,657 search results, 360 were reviewed, then combined with 43 publications that seeded the review and 1 obtained by citation tracking. 27 were included, published 1997–2022. The specificity of search terms was poor, and inter-rater reliability was low. Almost all models were predictive, the most common tasks being prognosis and diagnosis. Most studies used clinical, occasionally laboratory and image, data. Several addressed memory and runtime costs. A general technique that specializes to most of those reviewed involved matching, retrieval, fitting, and evaluation steps that could optionally be supervised, optimized, or recursively performed.

  Conclusions:
  Localized models have potential to improve the performance of clinical decision support tools while maintaining interpretability, but rigorous comparisons to competing methods must be conducted and computational hurdles must be overcome. We hope that our review will spur future work on efficiency, reproducibility, and user needs.

keywords:
  - case-based reasoning
  - localized modeling
  - nearest neighbors
  - clinical decision support
  - scoping review
bibliography:
  - SR-bibliography.bib
  - SR-reviewers.bib
  - SR-0-Found.bib
  - SR-2-Included.bib
  - SR-3-Tracked.bib
  - SR-5-Screened.bib
biblio-style: unsrt
output:
  rticles::arxiv_article:
    keep_tex: true
    latex_engine: xelatex
    # citation_package: natbib
    citation_package: "default"
header-includes: |
  \usepackage{lscape}
  \newcommand{\landscapebegin}{\begin{landscape}}
  \newcommand{\landscapeend}{\end{landscape}}
---

@@

A main driver of the development of clinical information systems (CIS) and clinical decision support (CDS) tools has been to leverage the advantages of routinely collected health data toward improving care.
These advantages include their immediate accessibility through the institutional EHR, billing database, or other source; their specificity to the institution and the population it serves; and the closeness in time of the available data.
Though by definition not collected for research use, the secondary research use of routinely collected health data has been put forth, as "practice-based evidence", a complement to the paradigm of evidence-based practice.

Over the same period of time, interest in individualizing care from population-derived evidence-based recommendations to specific patient needs has driven the application of advanced computational tools, including artificial intelligence (AI). Whereas classical models produced formulae involving a limited set of data elements that would be uniformly applied to all patients, AI models often process hundreds or more variables in opaque ways to yield predictions that depend on not only their values but their associations and interplay with each other. The loss of straightforward interpretations of these models has limited their practical uptake and motivated the construction of explanatory statistics for opaque models as well as the development of more interpretable complex models.

One of few methodologies to address all three of these established needs is one we call _localized modeling_, which appears to have been introduced several times independently under different names, and in slightly different forms.
The approach is a specialized form of case-based reasoning (CBR) that relies on a patient similarity measure[^psm] to extract a cohort of past or training cases that then inform the diagnosis, prognosis, or care of a new or test case.
Traditional CBR returns these retrieved cases to inform human decision-making; the related nearest neighbors (NN) technique generates predictions from cohorts automatically via averaging (regression) or voting (classification) of their outcomes.
In contrast, localized modeling fits families of predictive models, for example generalized linear models (GLMs), to retrieved similarity cohorts in order to generate predictions.
Localized models thus provide an individualized way to fit fully interpretable models to large population data.

@@

This approach harkens to the aspirational "green button" that would, in response to a query, automatically retrieve patient data from an institution’s records with which to conduct on-demand retrospective studies for an individual patient [@Longhurst2014].
Several hurdles face the deployment of such an approach in practice, but its benefits must first be demonstrated.
Our motivations in this scoping review are to describe the settings in and problems with which this approach has been tasked, to attempt to organize them within a common framework, and to evaluate the promise they show toward achieving this or other needs in clinical informatics.
Along the way, we attempt to reconcile terminology, summarize motivations and evaluations, and propose valuable follow-up work.

@@

@Kolodner1992 distinguished two styles of CBR: problem-solving, which is more procedural and used when objectives are more clearly defined, and interpretative, which provides categorizations and justifications for possible solutions. A similar dichotomy is commonly used to distinguish the performance criterion for the usefulness of predictive models from the interpretability criterion for their usability. In these terms, localized models are highly procedural: The step of fitting a predictive model to a retrieved cohort is an automated adaptive strategy [@Begum2011], and the parameters that govern cohort retrieval can be tuned alongside model hyperparameters in a machine learning (ML) workflow. Accordingly, localized models have been primarily used for and evaluated on their ability to predict classes or outcomes.
Nevertheless, as some recent applications have shown, they can be used to draw inferences about the importance of different risk factors to specific individuals.
While most ML models come equipped with measures of feature importance and model-agnostic tools can generate explanations for model predictions, these quantifications are not directly interpretable model components analogous to the split nodes of a decision tree (DT) or the coefficients of a GLM.
While the patient similarity measure used to retrieve each cohort may be complicated, the cohort itself can be directly inspected by the user. Provided the model family fitted to the cohorts is interpretable, the localized model inherits this property. Thus, localized models may embody both styles of CBR.

@@

Our goals in this review are (1) to describe applications of similarity-based localized models using patient-level health data and (2) to provide a general framework for the design and evaluation of such localized models.
We focus narrowly on localized models, rather than broadly on patient similarity--based clinical decision support tools, so that we may thoroughly assess their value in terms of reported evaluations and comparisons to other methods.
This also allows us to devise a high-level framework that specializes to the majority of these methods, which we then use to frame our discussion and recommendations.

@@

We expected the general framework to come down to three choices: A patient similarity measure, a cohort selection process, and a statistical model family. A general implementation based on this method would allow researchers to expedite every step of the analysis process, including retrieval, optimization, and evaluation, and enable sensitivity, robustness, and multiverse analyses that help identify the most consequential choices along the way.

@@

We repeated the search on 2024 January 24 on the PubMed and Web of Science databases to bring the results up to date.

@@

Within Covidence, we screened titles and abstracts for relevance. We decided on four eligibility criteria to expedite the screening process.
Articles must be written in English, for readability; they must be original studies, to exclude secondary sources with duplicate information; their use settings must be medical, clinical, or related, to keep our review topical; and it must not be clear that their use of our search terms was different from our intended meaning.
Each of two authors (AC and JCB) screened roughly half of the entries. They regularly reviewed each other’s decisions to improve consistency. In cases of uncertainty, entries were included.
For the update, one author (PM) screened all results for three criteria and another (JCB) screened the survivors for the fourth criterion.

@@

Finally, one author (JCB) applied the same review process to the seed set of 43 studies that inspired the review.
The same author later reviewed the updated results.

@@

We next collected characteristics of included studies. The features included bibliographic fields (date of publication, journal, authors, title, keywords, DOI), study goals (objective, generalizable knowledge, evaluation, clinical/medical domain), data sets (data source, type of data, range of data, number of cases/incidences, number of predictors/features), and methodological choices (types of similarity measure, families of adaptation step/statistical model, method(s) compared against, performance measures, results of evaluations and comparisons, name given to modeling approach). We used these data to detect and visualize trends amongst the included studies.

@@

Rather than an evidence synthesis characteristic of most systematic reviews, we here pursue a methodology synthesis to harmonize largely independent research efforts that have converged on a common family of techniques. The goal will be to describe a unified framework for localized models that can be used to guide future study designs and implementations as well as more systematically evaluate variations on the theme and measure the dependence of results on modeling choices.

@@

Figure \ref{fig:prisma} depicts our identification of studies via databases and registers. Following the completion and input of each search string, there were a total of 1,817 sources within all of the platforms used. De-duplication resulted in 1,657 entries, which were added to the title/abstract screening for review. Of these, 360 entries met the screening criteria and were assessed through full text review. Of these, 51 fit the original criteria, and 21 were included as distinct from NN prediction.

```{r prisma, fig.align='center', out.width="80%", echo=FALSE}
#| fig.cap =
#| "PRISMA-S flow chart.
#|  Lowercase letters refer to items obtained from
#|  the seed set (m),
#|  the structured search (n), and
#|  citation/reference tracking (s).
#|  Within the structured search results,
#|  summands correspond to original and update."
knitr::include_graphics("Fig1.pdf")
```

@@

We then reviewed 43 studies comprising a seed set that inspired this review. After removing duplicates and screening for eligibility, we were left with 6 additional studies [@Park2006; @Lowsky2013; @Lee2015; @Ng2015; @Lee2017; @Wang2019], 1 of which was excluded from the synthesis for using NN prediction. Reference tracking from the 43 + 360 = 403 studies assessed for eligibility led us to identify 1 additional study that met our criteria [@Kasabov2010]. This left us with 21 + 1 + 5 = 27 studies included in the review and synthesis.

@@

The `r nrow(tab_synthesis)` included studies were analyzed based on several characteristics, and we report and describe some observations here (Table \ref{tab:synthesis}). The studies were published in a variety of journals, with some of greater frequency, though no single journal published more than 3. The journal _Evolving Systems_ published 2 of the included studies, _Artificial Intelligence in Medicine_ published 3, _Hindawi - Journal of Healthcare Engineering_ published 3, \emph{Evolving Systems} published 2, and _PLOS One_ published 2.

Another interesting pattern lay in the years of publication (Figure \ref{fig:year}). While these studies trace back to the late 1990s, they have become more common, which suggests that this is an active, though not rapidly expanding, approach. We note that CBR in health and medicine originated as early as 1990, and localized modeling emerged soon after CBR had established itself; the idea has been "in the air" for as long as CBR has been in use.

@@

In addition to the aim of its analysis, we coded several aspects of the design of each study, including the source and type of data and the clinical task the model performed (Table \ref{tab:synthesis}), as well as the specific methodology and terminology adopted (Table \ref{tab:composite}).
In the following subsections, we take a closer look at these design elements and their reported justifications and limitations.

@@

The most common clinical motivations were individualized detection or diagnosis, prognosis or outcome prediction, and treatment or care recommendation. The plurality focused on prognosis or outcome prediction, often using time-to-event analysis: @Mariuzzi1997 used CBR to predict survival time from several geometric properties of breast tumors. @Lowsky2013 used CBR with non-parametric survival models on registry data to predict patient–graft survival times following kidney transplantation. @Lee2015 and @Lee2017 used localized logistic regression and random forest modeling to predict 30-day mortality following discharge for ICU patients. @Vilhena2016 developed a CBR cycle around a clustering-informed similarity matching procedure and an artificial neural network–based classifier to identify thrombophilia patients at high risk of thrombotic episodes. @Ma2020 took a similarity cohort–based approach to predicting length of stay for ICU patients. @Doborjeh2022 localized spiking neural networks in order to assess stroke risk from up to 7-day time series of clinical and environmental factors. @Liu2022 incorporated global-to-local transfer learning into localized models of acute kidney injury risk in hospitalized patients. Also of note, from a public health perspective, @Xu2008 used CBR to predict rehabilitation time as well as disability risk for unemployed workers experiencing chronic pain.

@@

The other frequent advantage attributed to localized modeling was interpretability.
@Elter2007 and @Nicolas2014 emphasized the value of the detailed past cases, available to the user, on which predictions are based. This "self-explanation capability" made outputs more intelligible to physicians in the CDS setting.
@Ng2015 additionally pointed out that localized GLMs yield localized effect estimates, which may help investigators identify individually relevant risk factors.
@Liu2022 took this idea further, systematically comparing regression coefficients for specific risk factors across dozens of diagnostic subpopulations.
@Doborjeh2022 used localized time series models to identify environmental changes associated with increased stroke risk.

@@

CBR can be understood as the opposite side of a trade-off with rule-based reasoning between model size and model complexity: @Doyle2006 describe their approach as "knowledge-light", in that "the cases do not contain explicit explanation structures; instead, explanation is achieved by comparison of the query case with retrieved cases". This means that the greatest performance and efficiency challenges in CBR have to do with the retrieval and revision phases in the R4 cycle. Several studies addressed these challenges: @Park2006, while excluded from the synthesis, were the earliest to propose that cohorts be bounded by a similarity threshold rather than by the number of cases, which improved performance in their experiments. @CampilloGimenez2013 leveraged predictor weights obtained from logistic regression to inform the similarity calculation used in retrieval. @Lowsky2013 used non-parametric models to reduce the computational burden of revision (prediction). @Ma2020 proposed to improve efficiency along the entire R4 cycle, but in particular by only executing task-dependent steps in real time ("just-in-time learning", JITL), and @Ng2021 and @Tang2021 partitioned multiple phases in the R4 cycle into offline and online components, only the latter of which would be performed in real time as new data are received.
@Liu2022 used transfer learning from globally-fitted models to improve the efficiency of localized models.

Other studies addressed limitations of available tools. @Lopez2011 implemented a comprehensive CBR tool in response to the lack of general-purpose software, to enable coupling with other tools as well as to expedite development, experimentation, and uptake. Several other teams also set out to develop more generalizable and data-agnostic clinical support tools [@Elter2007; @Liang2015; @Ng2015; @Zhang2018].
However, more informatical and legal challenges to implementation, including interoperability of systems and data and regulations concerning security and privacy, were rarely addressed.
We discuss this further in Section \ref{feasibility}.

@@

Most included studies quantitatively compared the predictive performance of their proposed method(s) to one or more comparators. What we took to be the signature results are collated in Table \ref{tab:performance} in the Appendix. Note that we exercised some judgment in classifying methods as proposals and comparators, as in some cases all methods were original but only some showcased main ideas.
It would be impractical to meta-analyze these numbers due to the great variety of settings, problems, data types, techniques, and choices involved.

We do observe one clear pattern: All of the proposed methods that most evidently outperformed their comparators---Kohonen + CBR [@Wyns2004], TWNFI [@Song2006, @Kasabov2010], gravitational search algorithms (GSA) [@Liang2015], CBR + rules (+ DML) [@Nicolas2014], Gaussian process regression [@Zhang2018], JITL-ELM [@Ma2020]---are hybrids of localized modeling (in some cases CBR) with other techniques, often DML. Though @CampilloGimenez2013 and @Ng2015 report non-superior performance by such hybrids, the pattern suggests the importance of the similarity measure to the retrieval step.
Meanwhile, when proposed approaches targeted cohort demarcation or choice of predictive model---statistical CBR [@Park2006]; individualized logistic regression, decision tree, and random forest [@Lee2015; @Lee2017]---they did not consistently outperform comparators.

@@

Consistent with their orientation, almost all included studies were conducted using clinical data, only occasionally together with patient-reported (2), laboratory (3), image (1), and environmental (1) data. We suggest three reasons for this: First, this literature traces back to the 1990s, before -omic data could be generated cost-effectively at scale. Second, CBR in particular has a strong tradition in clinical decision support, where the focus of our sample remains throughout the review period. Third, because most -omic data are highly homogeneous—all measurements are made along or can be transformed to a common scale, e.g. greyscale pixellations for X-ray images and transcripts per million for RNA-seq data—more deeply theoretical analysis techniques have been developed and come into wide use. While variations on correlation-based approaches like EHR-based phenome-wide association studies and similarity-based methods like CBR itself have been developed, more mechanistic and probabilistic tools have not become a domain standard.

@@

```{r framework, fig.align='center', out.width="100%", echo=FALSE}
#| fig.cap =
#| "General framework for localized models.
#|  Dashed arrows indicate optional steps."
knitr::include_graphics("Fig5.png")
```
